### 2020-12-09 doc/manuscript/Dow_manuscript.Rmd
intro work
rearranging for clarity / more focus on stem-phenology instead of leaf phenology

climwin_caption <- "Figure 3. Landscapes of relationships between the day of year on which 25% of annual growth is achieved (DOY_25) and temperature in prior weeks for ring- and diffuse-porous trees at SCBI and Harvard Forest. Shown are matrices of linear coefficients of first-order linear regressions between temperature and DOY_25, where Window Open and Window Close indicate number of weeks prior to DOY_25 (ring-porous: May 4 at SCBI, May 15 at HF; diffuse-porous: June 4 at SCBI, June 14 at HF). Black boxes indicate the optimal window (ring-porous:  March 22- April 9 at SCBI, April 2 - May 07 at HF; diffuse-porous: Feb. 19- May 21 at SCBI, March 19 - May 07 at HF)."

pheno_Tsensitivity_caption <- "Figure 4. Relationship between growth parameters and climwin mean temperatures. For each 
observed climwin mean temperature value, the posterior mean of the fitted day of year
is represented by the solid blue line and credible intervals are represented by bands."

DOYtiming_caption <- "Figure 5. Day of year where growth milestones were achied at (a) the Smithonian Conservation Biology Institute and (b) Harvard Forest."

tree_rings_caption <- "Figure 6. Climate sensitiivity of annual growth, as derived from tree-rings, for dominant ring- and diffuse- porous species at the Smithonian Conservation Biology Institute and Harvard Forest. Shown are Pearson correlations between ring-width index and and monthly Tmax. Species codes are as given in Table 2."

@@

#SI figure captions
climwin_rp_DOY25_SCBI_short_caption <- "Figure S1. Full climwin output for DOY 25 for ring porous species at SCBI."
climwin_dp_DOY25_SCBI_short_caption <- "Figure S2. Full climwin output for DOY 25 for diffuse porous species at SCBI."
climwin_rp_DOY25_HARV_short_caption <- "Figure S3. Full climwin output for DOY 25 for ring porous species at Harvard Forest."
climwin_dp_DOY25_HARV_short_caption <- "Figure S4. Full climwin output for DOY 25 for diffuse porous species at Harvard Forest."
tree_rings_TMIN_short_caption <- "Figure S5. ..."

@@

Reducing this uncertainty requires better understanding of the drivers of tree growth (Bugmann et al, 2019). Tree growth is sensitive to climatic factors like precipitation, temperature, and cloudiness in early growing season (Helcoski et al, 2019; Ciais, 2003; Goulden, 1996), as well as other physical factors including nitrogen content of soil, increasing atmospheric CO~2~, deposition of atmospheric pollutants in acid rain (Mathias & Thomas, 2018), and invasion by pests or pathogens. Understanding how growth is influenced by these factors is essential for developing accurate models to inform forest managers and policy makers about the role of future forests in climate change mitigation efforts.

## (spring temperatures are rising, and this is affecting leaf phenology and carbon cycling)
In recent decades, an increase in average spring temperature has been impacting many natural systems including forests (Parmesan & Yohe, 2003; Zhihong, 2006). 
Shifts in leaf phenology timing could be accompanied by changes in stem-growth phenology, which would likely affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests (Keeling et al, 1996;). 
The effects of this spring warming are most apparent when examining leaf phenology shifts, *which have been extensively studied (Way, 2011; Klosterman & Hufkins & Richardson, 2018), leading to many critical observations*. As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, sometimes matched or exceeded by later leaf senescence, meaning a potential shift in growing season timing or length (Ibáñez et al, 2010; Jeong et al., 2011). The timing of this spring warming also has a significant effect on leaf phenology response (Crabbe et al, 2016).

Leaf phenology is linked to the ecosystem CO~2~ uptake period, and in turn longer carbon uptake periods have been linked to greater net annual C sequestration (*i.e.*, net ecosystem exchange, $NEE$; @churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016]. 
However, more of the interannual variation in forest $NEE$ is explained by variation in maximum C uptake rates than by the length of the carbon uptake period [@zhou_explaining_2016; @fu_maximum_2019].

Thus, while it is increasingly clear that warming spring temperatures lengthen the growing season and modestly increase annual carbon sequestration, we know little about the effect on tree growth and woody productivity, which can be decoupled from leaf phenology (D'Orangeville et al, in review) and represent only a modest proportion of total C sequestration (Anderson-Teixeira et al. in review).

## (However, we don't know how warming springs affect stem-growth, which matters because the C allocated to stem growth will be retained in the ecosystem for decades to millenia )

Earlier timing of spring leaf-up and initiation of carbon sequestration could signal that these trees are shifting their growth windows in response to increasing spring temperatures, but to date, very little research has been done on this topic. 
In fact, we know little about the stem-growth phenology of temperate deciduous species — let alone how these are affected by changes in spring temperature patterns. *(but check into this.... I suspect there may be some that we're missing--perhaps auto dendro bands in Europe. Neil did say that there's little on early spring T)*
Warming events in late-spring were shown to shift leaf emergence timing back more than early-spring warming (Freidl et al, 2014) and earlier leaf emergence has been shown to cause complex lagged effects on subsequent seasonal growth in European broad leaf deciduous and mixed forests (Crabbe et al, 2016). Identifying the pre-season period of greatest sensitivity of intraannual growth, along with the effect warming within that period has on other stem phenology traits could help unravel the mysteries of tree growth.

Tree-ring records give some hints that warmer springs may increase annual growth (Helcoski et al. 2019; probably some more), but tree-rings tell us nothing about the timing of that growth.  Timing of intraannual stem-growth is extremely important as well.

## (We also know little about differences across species or functional groups, which matters because it will affect how different species fare under climate change )
Leaf and wood phenology  traits, along with their apparent relationships, are dissimilar among trees of different wood types (ring-porous and diffuse-porous; **D'Orangeville et al. **). For example, bud burst occurs around the same time as 25% total annual growth, and 75% leaf development coincided with 50% total annual growth in ring-porous trees, while leaf phenology and stem growth phenology were decoupled in diffuse-porous trees.

The significance of wood type in regards to wood and leaf phenology likely stems from evolutionary differences and adaptations to local climates where these trees can be found. Ring-porous trees have large xylem cells which are prone to embolism during the winter. These cells are repaired before leaf development, explaining why the 25% yearly growth milestone is reached around the same time as bud burst. Diffuse-porous xylem cells are smaller, and less prone to embolism during the Winter, meaning they do not need to be repaired in order to restore hydraulic functionality before beginning the leaf development process. Results from the paper by D’Orangeville et al. (in review) have shown that ring-porous trees reach 25 and 50% of their total yearly growth during spring, while diffuse-porous trees reach these milestones in the summer, but both reached 75% total annual growth around the same time. If leaf development is sped up when spring temperatures are warmer, we should seek to understand whether this relationship holds, and how it potentially varies among trees of different wood-types.

## (Here, we...)
Here, we characterize how early spring temperatures affect stem growth phenology and rates of temperate deciduous trees within two forest dynamics plots in the Eastern USA. Using data from dendrometer bands measured throughout the growing season, we fit a growth model  to the time series of individual trees to determine the day of year (DOY) where 25, 50, and 75% annual growth was achieved; maximum growth rates and the DOY on which they occurred; total annual growth; and peak growing season length (75%-25% DOY) [Fig. `r schematic`; @mcmahon_general_2015]. We test the hypotheses that (1) warmer early springs result in earlier stem growth and a period of growth, but (2) maximum growth rates are independent of spring temperatures, and as a result (3) total annual growth increases in response to warmer spring temperatures (Table `r hypothesis_table`).

@@

Our first study site was the forest dynamics plot located at the Smithsonian Conservation Biology Institute 5 km South of Front Royal, Virginia within the Blue Ridge Mountains (38.8935° N, 78.1454° W; elevation 273–338 m.a.s.l.). 
The forest here was established in the mid-19th century after conversion from agricultural fields. 
Dominate canopy species within the 25.6 ha forest plot include tulip poplar (*Liriodendron tulipifera*), oaks (*Quercus spp.*), hickories (*Carya spp.*). 
The climate is humid temperate, with a mean annual precipitation of 998 mm and temperatures averaging 1° C in January and 24° C in July (Helcoski et al, 2019).

@@

Although the number of years analyzed here is limited (# years at SCBI, 6 years at HF), the signal is clear and consistent: for both species groups at both sites, warmer temperatures during virtually any part of the spring led to significantly earlier achievement of growth milestones (Table `r hypothesis_table`, Figs. `r  climwin`-`r DOYtiming`).
This indicates that the effect of $T_{max}$ on the timing of growth ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$,and $DOY_{g_{max}}$) is not subtle; 
indeed, growth milestones advanced by #-# days per degree Celsius during the critical temperature window.
This is consistent with previous studies showing that ... *(find/cite some statistics on how rapidly leaf phenology or start of C sequestration advance with T)*

>>>>

climwin_caption <- "Figure 3. Landscapes of relationships between the day of year on which 25% of annual growth is achieved ($DOY_{25}$) and temperature in prior weeks for ring- and diffuse-porous trees at SCBI and Harvard Forest. Shown are matrices of linear coefficients of first-order linear regressions between temperature and $DOY_{25}$, where Window Open and Window Close indicate number of weeks prior to $DOY_{25}$ (ring-porous: May 5 at SCBI, May 13 at HF; diffuse-porous: June 4 at SCBI, June 14 at HF). Black circles indicate the critical $T_{max}$ window (ring-porous:  March 22- April 9 at SCBI, April 2 - May 07 at HF; diffuse-porous: Feb. 19- May 21 at SCBI, March 19 - May 07 at HF)."

pheno_Tsensitivity_caption <- "Figure 4. Relationship between growth parameters and climwin mean temperatures. For each 
observed climwin mean temperature value, the posterior mean of the fitted day of year
is represented by the solid blue line and 95% credible intervals are represented by bands."

DOYtiming_caption <- "Figure 5. Day of year where growth milestones were achieved at (a) the Smithonian Conservation Biology Institute and (b) Harvard Forest. Each line represents a single year of ring-porous (dotted line) or diffuse-porous (solid line) trees at each site. Mean temperature was calculated for each wood-type/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red."

tree_rings_caption <- "Figure 6. Climate sensitivity of annual growth, as derived from tree-rings, for dominant ring- and diffuse- porous species at the Smithsonian Conservation Biology Institute and Harvard Forest. Shown are Pearson correlations between ring-width index and and monthly Tmax. Species codes are as given in Table 2."

@@

#SI figure captions
climwin_rp_DOY25_SCBI_short_caption <- "Figure S1. Full climwin output for $DOY_{25}$ for ring porous species at SCBI."
climwin_dp_DOY25_SCBI_short_caption <- "Figure S2. Full climwin output for $DOY_{25}$ for diffuse porous species at SCBI."
climwin_rp_DOY25_HARV_short_caption <- "Figure S3. Full climwin output for $DOY_{25}$ for ring porous species at Harvard Forest."
climwin_dp_DOY25_HARV_short_caption <- "Figure S4. Full climwin output for $DOY_{25}$ for diffuse porous species at Harvard Forest."
tree_rings_TMIN_short_caption <- "Figure S5. ..."

@@

Reducing this uncertainty requires better understanding of the drivers of tree growth [@bugmann_tree_2019]. Tree growth is sensitive to climatic factors like precipitation, temperature, and cloudiness in early growing season [@Helcoski_growing_2019]; [@Ciais_europe-wide_2005]; [@Goulden_exchange_1996], as well as other physical factors including nitrogen content of soil, increasing atmospheric CO~2~, deposition of atmospheric pollutants in acid rain [@mathias_disentangling_2018], and invasion by pests or pathogens. Understanding how growth is influenced by these factors is essential for developing accurate models to inform forest managers and policy makers about the role of future forests in climate change mitigation efforts.

## (spring temperatures are rising, and this is affecting leaf phenology and carbon cycling)
In recent decades, an increase in average spring temperature has been impacting many natural systems including forests ([@parmesan_globally_2003]; [@zhizong_fingerprinting_2006]). 
The effects of this spring warming are most apparent when examining leaf phenology shifts, *which have been extensively studied ([@way_tree_2011]; [@klosterman_later_2018]), leading to many critical observations*. For example, some regions have seen earlier leaf out in deciduous trees, which is often matched or exceeded by delays in leaf senescence timing, when springtime temperatures are above average [@ibanez_forecasting_2010];@jeong_phenology_2011];@crabbe_extreme_2016]. The emergence of leaves signals the beginning of carbon sequestration or in other words, the growing season of a tree. It is believed that the longer the tree has leaves - or as the growing season is extended - the net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$; @churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016].)

However, more of the interannual variation in forest $NEE$ is explained by variation in maximum C uptake rates than by the length of the carbon uptake period [@zhou_explaining_2016; @fu_maximum_2019]. This implies that earlier leafout will lead to smaller increases in NEE than changes in other phenological traits like the maximum C uptake rate. This idea agrees with patterns seen in tree cores, where peak growing season climate was more influential than spring climate [@Helcoski_growing_2019] but many models still predict the extension of growing season will increase NEE. A paper by @zani_increased_2020 found that earlier leaf out led to earlier leaf senescence within the deciduous trees in their study, reducing the ability of the trees to sequester significantly higher levels of C. Instead they posit that senescence is induced when seasonal carbon-sink capacity of the tree is reached, no matter how extended the potential growing season is. This result directly contradicts those previous predictions that warmer Spring temperatures and earlier leaf out will lead to drastically increased C sequestration.

Thus, while it is increasingly clear that warming spring temperatures lengthen the growing season and modestly increase annual carbon sequestration, indirect observations of these effects - like those involving leaf phenology - are not enough to draw firm conclusions on the future of forests. Instead, direct observations of intraannual stem-growth should be compared between years with a variety of spring climate conditions so a clearer relationship between climate and stem-growth can be deciphered. Looking directly at the stem-growth is advantageous as it allows examination of individual trees as opposed to larger landscape scale patterns obtained through eddy-covariance or remote sensing studies. A direct, fine scale, individual based study would allow a look into the effect of warming springtime on stem-growth phenology and the related shifts in growing season length and $NEE$.

## (However, we don't know how warming springs affect stem-growth, which matters because the C allocated to stem growth will be retained in the ecosystem for decades to millenia )
Earlier timing of spring leaf-up and initiation of carbon sequestration signals that these trees are shifting their growth windows in response to increasing spring temperatures, but to date, very little research on the characteristics of stem-growth during this shift has been done. 
In fact, we know little about the stem-growth phenology of temperate deciduous species — let alone how these are affected by changes in spring temperature patterns. *(but check into this.... I suspect there may be some that we're missing--perhaps auto dendro bands in Europe. Neil did say that there's little on early spring T)*.

Tree-ring records give some hints that warmer springs may increase annual growth ([@Helcoski_growing_2019]; probably some more), but tree-rings tell us nothing about the timing of that growth. Similarly, large scale observations of leaf-phenolgy and eddy-covariance also hint that warmer springs increase annual growth but tell us little about the maximum C uptake rates, or interspecific differences in total carbon sequestered.

## (We also know little about differences across species or functional groups, which matters because it will affect how different species fare under climate change )
Leaf and wood phenology traits, along with their apparent relationships, are dissimilar among trees of different wood types (ring-porous and diffuse-porous; **D'Orangeville et al. **). Interspecific variation of this sort likely stems from evolutionary differences and adaptations to local climates where these trees can be found. Ring-porous trees have large xylem cells which are prone to embolism during the winter. These cells are repaired before leaf development in the Spring, corresponding with an earlier initiaiton of stem-growth. Diffuse-porous xylem cells are smaller, and less prone to embolism during the Winter, meaning they do not need to be repaired in order to restore hydraulic functionality before beginning the leaf development process. Results from the paper by D’Orangeville et al. (in review) have shown that ring-porous trees reach 25 and 50% of their total yearly growth during spring, while diffuse-porous trees reach these milestones in the summer, but both reached 75% total annual growth around the same time. If leaf development is sped up when spring temperatures are warmer, we should seek to understand how this shift influences stem-growth timing and other phenological variables.

## (Here, we...)
Here, we characterize how early spring temperatures affect stem growth phenology and rates of temperate deciduous trees within two forest dynamics plots in the Eastern USA. Using data from dendrometer bands measured throughout the growing season, we fit a growth model to the time series of individual trees to determine the day of year where 25, 50, and 75% annual growth was achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$); maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$); total annual growth ($\Delta DBH$); and peak growing season length ($DOY_{75}$-$DOY_{25}$; $L_{pgs}$) [Fig. `r schematic`; @mcmahon_general_2015]. We test the hypotheses that (1) warmer early springs result in earlier stem growth and a longer period of growth, but (2) maximum growth rates are independent of spring temperatures, and as a result (3) total annual growth increases in response to warmer spring temperatures (Table `r hypothesis_table`).

@@

Our first study site was the forest dynamics plot located at the Smithsonian Conservation Biology Institute 5 km South of Front Royal, Virginia within the Blue Ridge Mountains (38.8935° N, 78.1454° W; elevation 273–338 m.a.s.l.). 
The forest here was established in the mid-19th century after conversion from agricultural fields. 
Dominate canopy species within the 25.6 ha forest plot include tulip poplar (*Liriodendron tulipifera*), oaks (*Quercus spp.*), hickories (*Carya spp.*). 
The climate is humid temperate, with a mean annual precipitation of 998 mm and temperatures averaging 1° C in January and 24° C in July ([@Helcoski_growing_2019]).

@@

Although the number of years analyzed here is limited (10 years at SCBI, 6 years at HF), the signal is clear and consistent: for both species groups at both sites, warmer temperatures during virtually any part of the spring led to significantly earlier achievement of growth milestones (Table `r hypothesis_table`, Figs. `r  climwin`-`r DOYtiming`).
This indicates that the effect of $T_{max}$ on the timing of growth ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$,and $DOY_{g_{max}}$) is not subtle; 
indeed, growth milestones advanced by #-# days per degree Celsius during the critical temperature window.
This is consistent with previous studies showing that ... *(find/cite some statistics on how rapidly leaf phenology or start of C sequestration advance with T)*
